A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-10-08 introProtocol.tex
Completed the section of Schedule Reset

CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{d}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.

@@

In Figure~\ref{fig:BECA-example}, the \emph{STA-\#} labels represent stations willing to transmit. The horizontal lines represent a time axis with each number indicating the amount of empty slots left for the backoff to expire. Stations willing to transmit begin the contention for the channel by waiting a random backoff, $B$. The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$. After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{d}=7$ in this case. By doing so, all successful STAs will not collide among each other in future cycles.

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA} (redrawn from~\cite{research2standards}), shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.

\begin{figure}[tb]
\centering
  \includegraphics[width=0.7\linewidth, angle=-90]{figures/DCF-vs-ECA/DCF-v-ECA-TON.eps}
  \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
  \label{fig:BECA}
\end{figure}

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{d}$ ($B_{d}=7$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's.

The \emph{JFI for CSMA/ECA} and \emph{JFI for CSMA/CA} curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for both protocols. Both protocols have a JFI equal to one, which suggests that the available throughput is shared evenly among all stations.

	\subsection{Supporting many more contenders}\label{moreContenders}
	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{d}$. When $N > B_{d}$, collisions reappear.

	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{d}$, we introduce \emph{Hysteresis}. Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{d}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$. This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule. This idea of a schedule is significantly different from the virtual schedule required by the protocols described in Section~\ref{relatedWork}, that is, CSMA/ECA with Hysteresis does not require a previous knowledge of the number of contenders, the result of previous transmissions or the start/end of the schedule, easing its implementation in real hardware.

	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{d}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).

@@

	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{d}=CW(k)/2$. The same behavior is followed by STA 3.

	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown in Figure~\ref{fig:ECA+H+F-throughput} extracted from~\cite{research2standards}. In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission. This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming.

	\begin{algorithm}[tb]
	\While{the device is on}
	{
	  $r \leftarrow 0$ ; $k \leftarrow 0$\;
	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
	  \While{there is a packet to transmit}{
	    %\tcc{Initialize $a$.}
	    \Repeat{($r = R$) or (success)}{
	      %\tcc{First, backoff.}
	      \While{$B>0$}{
	        wait 1 slot\;
	        $B \leftarrow B-1$\;
	      }
	      \colorbox{yellow}{Attempt transmission of $2^k$ packets;}\\
	      \If{collision}{
	        %\tcc{Random backoff.}
	        $r \leftarrow r+1$\;
	        $k \leftarrow \min (k+1,m)$\;
	        $B \leftarrow \mathcal{U}[0, 2^k  \rm{CW}_{min} -1]$\;
	      }
	    }
	    $r \leftarrow 0$\;
	    %$s \leftarrow 0$\;
	    \eIf{success}{
	      %\tcc{Random backoff.}
	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\\
		$B \leftarrow B_{d}$;\\
	    }
	    {
	      Discard packet\;
	      $k \leftarrow 0$\;
	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
	    }
	  }
	  Wait until there is a packet to transmit\;
	}
	\vspace{0.2cm}
	\caption{CSMA/ECA$_{\text{Hys+FS}}$}
	\label{alg:fullECA}
	\end{algorithm}

@@

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an \textcolor{red}{A-MPDU} aggregation mechanism that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video. \textcolor{red}{An example of the duration of a transmission is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}.}

	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/CA$_{\text{MaxAg}}$ and CSMA/ECA$_{\text{Hys+MaxAg}}$. Although CSMA/CA$_{\text{MaxAg}}$ performs maximum aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission. On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg-TON.eps}
		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$. \textcolor{red}{This figure should also show CSMA/CA$_{\text{FS}}$}.}
		\label{fig:ECA-vs-DCF-maxAgg}
	\end{figure}

@@

	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share. The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation. That is, with a schedule length of $C=2^{k}B_{d}$, where $k = \left\lceil log_{2}(N/B_{d})\right\rceil$. Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{d}$. Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{red}{nodes the $(k-1)$-th one}. The system throughput is computed as follows:

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{d} \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

@@

%	\textcolor{red}{\subsection{Real hardware implementation} 
%	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure. In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
%	
	\subsection{Backwards compatibilty and coexistence}
	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism. It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network. Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_d$, is the expected value for the current backoff stage $k$ ($B_d\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders. An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

>>>>

CSMA/ECA~\cite{barcelo2008lba} is a fully distributed and collision-free MAC for WLANs. It differs from CSMA/CA in that it uses a deterministic backoff, $B_{\text{d}}=CW_{\min}/2$ after successful transmissions, where $CW_{\min}$ is the minimum contention window of typical value $CW_{\min}=16$. By doing so, contenders that successfully transmitted on schedule $n$, will transmit without colliding with other successful nodes in future cycles.

@@

In Figure~\ref{fig:BECA-example}, the \emph{STA-\#} labels represent stations willing to transmit. The horizontal lines represent a time axis with each number indicating the amount of empty slots left for the backoff to expire. Stations willing to transmit begin the contention for the channel by waiting a random backoff, $B$. The first outline highlights the fact that stations STA-3 and STA-4 will eventually collide because they have selected the same $B$. After recomputing the random backoff, STA-4's attempt results in a successful transmission, which instructs the node to use a deterministic backoff, $B_{\text{d}}=7$ in this case. By doing so, all successful STAs will not collide among each other in future cycles.

Collision slots being orders of magnitude larger than empty slots degrade the network performance. When CSMA/ECA builds the collision-free schedule all contenders are able to successfully transmit more often, increasing the aggregated throughput beyond CSMA/CA's. Figure~\ref{fig:BECA}, \textcolor{red}{shows the average aggregated throughput of CSMA/ECA, CSMA/CA and two of the collision-free protocolos presented in Section~\ref{relatedWork}, namely L-ZC and L-MAC~\cite{L_MAC}}.

%shows the achieved throughput of CSMA/ECA and CSMA/CA, alongside the Jain's Fairness Index (JFI)~\cite{JFI}.

\begin{figure}[tb]
\centering
  \includegraphics[width=0.7\linewidth, angle=-90]{figures/tonFigs/DCF-v-ECA-TON.eps}
  \caption{CSMA/ECA example in saturation: throughput (simulation parameters can be found in Section~\ref{simulations})}
  \label{fig:BECA}
\end{figure}

Referring to Figure~\ref{fig:BECA}, CSMA/ECA is able to achieve an aggregated throughput that goes beyond CSMA/CA up until the number of contenders ($N$) is greater than $B_{\text{d}}$ ($B_{\text{d}}=7$ in the case of the figure). Beyond this point, the network will have a mixed behavior relating to backoff mechanisms: some nodes will successfully transmit and use a deterministic backoff while others will collide due to the lack of empty slots and return to a random backoff. As more contenders join the network, CSMA/ECA performance will approximate to CSMA/CA's. \textcolor{red}{This effect is also observed in L-ZC and L-MAC, albeit at a larger number of nodes. This is becasue the virtual cicle used by these protocolos ($N$ for L-ZC and $C$ for L-MAC in Section~\ref{relatedWork}) in greater than B$_{d}$. Specifically, the virtual cycle is of size equal to $16$.}

The \emph{JFI } curves in Figure~\ref{fig:BECA} show the Jain's Fairness index for all tested protocols. Showing a JFI equal to one suggests that the available throughput is shared evenly among all stations.

	\subsection{Supporting many more contenders}\label{moreContenders}
	As was mentioned before, CSMA/ECA is only able to build a collision-free schedule if the number of contenders $N$, is less or equal than $B_{\text{d}}$. When $N > B_{\text{d}}$, collisions reappear.

	To be able to attain a collision-free schedule even when the number of contenders exceeds $B_{\text{d}}$, we introduce \emph{Hysteresis}. Hysteresis is a property of the protocol that instructs nodes not to reset their backoff stage ($k$) after successful transmissions, but to use a deterministic backoff $B_{\text{d}}=CW(k)/2$, where $CW(k)=2^{k}CW_{\min}$. This measure allows the adaptation of the schedule length, admitting many more contenders in a collision-free schedule. This idea of a schedule is significantly different from the virtual schedule required by the protocols described in Section~\ref{relatedWork}, that is, CSMA/ECA with Hysteresis does not require a previous knowledge of the number of contenders, the result of previous transmissions or the start/end of the schedule, easing its implementation in real hardware.

	Hysteresis enables CSMA/ECA nodes to have different schedules ($B_{\text{d}}$), carrying the undesired effect of unevenly dividing the channel time among contenders (i.e., some nodes will have to wait more in order to attempt transmissions).

@@

	Algorithm~\ref{alg:fullECA} shows an implementation of CSMA/ECA$_{\text{Hys+FS}}$, while an example of CSMA/ECA$_{\text{Hys+FS}}$ with four contenders is shown in  Figure~\ref{fig:ECA+Hyst}. In the figure the first outline indicates a collision between STA-3 and STA-4, which will provoke an increment on both station's backoff stage ($k\leftarrow k+1$). Once STA-4's random backoff expires, CSMA/ECA$_{\text{Hys+FS}}$ instructs the station to transmit $2^{k}$ packets, and then use a deterministic backoff, $B_{\text{d}}=CW(k)/2$. The same behavior is followed by STA 3.

	With Hysteresis and Fair Share, CSMA/ECA$_{\text{Hys+FS}}$ is able to achieve greater throughput than CSMA/CA and for many more contenders, as shown in Figure~\ref{fig:ECA+H+F-throughput} extracted from~\cite{research2standards}. In the figure, the \emph{CSMA/ECA$_{\text{Hys+FS}}$} curve shows a greater throughput because collisions are eliminated and Fair Share allows nodes to send $2^{k}$ packets upon each transmission. This throughput increase is the result of aggregation via Fair Share, which also carries the effect of raising the average time between successful transmissions (see Section~\ref{timeBetweenSxTx}), which may affect delay-sensitive traffic, like gaming or live video/voice/tv streaming. \textcolor{red}{Channel errors or unsaturated traffic conditions will disrupt any collision-free schedule, generate collisions and push all stations's backoff stages to the maximum value very quickly (contributing to the delay increase). This issue is leveraged with Schedule Reset mechanims, presented in Section~\ref{schedReset}.}

	\begin{algorithm}[tb]
	\While{the device is on}
	{
	  $r \leftarrow 0$ ; $k \leftarrow 0$\;
	  $b \leftarrow \mathcal{U}[0,2^k\rm{CW}_{min}-1]$\;
	  \While{there is a packet to transmit}{
	    %\tcc{Initialize $a$.}
	    \Repeat{($r = R$) or (success)}{
	      %\tcc{First, backoff.}
	      \While{$B>0$}{
	        wait 1 slot\;
	        $B \leftarrow B-1$\;
	      }
	      \colorbox{yellow}{Attempt transmission of $2^k$ packets;}\\
	      \If{collision}{
	        %\tcc{Random backoff.}
	        $r \leftarrow r+1$\;
	        $k \leftarrow \min (k+1,m)$\;
	        $B \leftarrow \mathcal{U}[0, 2^k  \rm{CW}_{min} -1]$\;
	      }
	    }
	    $r \leftarrow 0$\;
	    %$s \leftarrow 0$\;
	    \eIf{success}{
	      %\tcc{Random backoff.}
	      \colorbox{yellow}{$B_{d} \leftarrow (2^{k}\rm{CW}_{min})/2-1$;}\label{backoffExample}\\
		$B \leftarrow B_{d}$;\\
	    }
	    {
	      Discard packet\;
	      $k \leftarrow 0$\; \label{labelReset}
	      $B \leftarrow \mathcal{U}[0, 2^k \rm{CW}_{min}-1]$\;
	    }
	  }
	  Wait until there is a packet to transmit\;
	}	
	\vspace{0.2cm}
	\caption{CSMA/ECA$_{\text{Hys+FS}}$}
	\label{alg:fullECA}
	\end{algorithm}

	\subsection{The effects of Aggregation}\label{effects-of-aggregation}
	Fair Share is an \textcolor{red}{A-MPDU} aggregation mechanism~\cite{A-MPDU} that coupled with the collision-free schedule built by CSMA/ECA$_{\text{Hys}}$ is able to provide short-term fairness. However, it also improves the throughput since the aggregation process makes the packet transmission more efficient by reducing overheads. The downside of Fair Share is that it may increase the time between two consecutive transmissions from the same node, which may affect negatively delay-sensitive applications such as gaming or high definition real-time video. \textcolor{red}{An example of the duration of a transmission, $T(l_{k})$, is provided by (\ref{eq:Ts}) in the following Section~\ref{ECA-bounds}.}

	Figure~\ref{fig:ECA-vs-DCF-maxAgg} shows the aggregated throughput for CSMA/ECA$_{\text{Hys+FS}}$, CSMA/ECA$_{\text{Hys+MaxAg}}$, CSMA/CA using the Fair Share mechanism (CSMA/CA$_{\text{FS}}$), and CSMA/CA$_{\text{MaxAg}}$. Although CSMA/CA$_{\text{FS}}$ and CSMA/CA$_{\text{MaxAg}}$ perform aggregation, collisions degrade the aggregated throughput as more contenders attempt transmission. On the other hand, CSMA/ECA$_{\text{Hys+FS}}$ is able to build a collision-free schedule and takes advantage of the aggregation provided by Fair Share, which opposed to just using CSMA/ECA$_{\text{Hys+MaxAg}}$, it is fair (see Figure~\ref{fig:fairness-with-maximum-aggregation}).

@@

	\begin{figure}[tb]
	\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/throughput-max-Ag/throughput-saturated-maxAgg-TON.eps}
		\caption{Throughput comparison with CSMA/CA$_{\text{MaxAg}}$: even though at low number of contenders CSMA/CA$_{\text{MaxAg}}$ achieves greater throughput than CSMA/ECA$_{\text{Hys+FS}}$, collisions eventually degrade the throughput below CSMA/ECA$_{\text{Hys+FS}}$'s when the number of contenders increases past $N=10$}
		\label{fig:ECA-vs-DCF-maxAgg}
	\end{figure}

@@

	They correspond to the maximum and minimum achievable throughput without the possibility of collisions using Hysteresis and Fair Share. The ideal CSMA/ECA$_{\text{Hys+FS}}$ network uses the minimum schedule length that guarantees a collision-free operation. That is, with a schedule length of $C=2^{k}B_{\text{d}}$, where $k = \left\lceil log_{2}(N/B_{\text{d}})\right\rceil$. Using this minimum schedule length, $N$ nodes will be at the same backoff stage if $N\leq B_{\text{d}}$. Otherwise, $h = N-(C-N)$ nodes would occupy the $k$-th backoff stage and the other $N-h$ \textcolor{red}{nodes the $(k-1)$-th one}. The system throughput is computed as follows:

	\begin{equation}
			S =
				\begin{cases}\label{eq:bound}
					hs_{k}(l_{k}) + (N-h)s_{k-1}(l_{k-1}), & \text{if } N > B_{\text{d}} \\
					Ns_{k}(l_{k}), & \text{otherwise}
				\end{cases}
	\end{equation}

@@

%	\textcolor{red}{\subsection{Real hardware implementation} 
%	Replacing the firmware of commercial WLAN cards with the Open Firmware for WiFi Networks, OpenFWWF~\cite{OpenFWWF}, allows the modification of time-sensitive operations in the MAC, like the backoff procedure. In Section~\ref{EDCA} we provide details regarding the tesbed, the firmware used and present results of real experiments using CSMA/ECA.}
%	
	\subsection{Backwards compatibilty and coexistence}
	CSMA/ECA$_{\text{Hys+FS}}$ springs from a modification to CSMA/CA's backoff mechanism. It keeps the range of values CSMA/CA nodes use to draw a random backoff (i.e., use the same $CW_{\min}$ and $CW_{\max}$), allowing CSMA/ECA$_{\text{Hys+FS}}$ contenders to coexist with CSMA/CA nodes in the same network. Further, the selection of CSMA/ECA$_{\text{Hys+FS}}$'s deterministic backoff, $B_{\text{d}}$, is the expected value for the current backoff stage $k$ ($B_{\text{d}}\coloneqq\lceil{E[0,CW(k)]}\rceil$)~\cite{research2standards}, which ensures fairness among contenders. An overview of the attained throughput for different proportions of CSMA/ECA$_{\text{Hys+FS}}$ and CSMA/CA nodes is presented in Section~\ref{coexistance-w-csmaca}.

	\textcolor{red}{\subsection{Schedule Reset Mechanism}\label{schedReset}}
	CSMA/ECA$_{\text{Hys+FS}}$ instructs nodes to keep their current backoff stage after a successful transmission (resetting it to zero only if reaching the retransmission attempt limit, see line~\ref{labelReset} in Algorithm~\ref{alg:fullECA}). This is done in order to increase the cycle length and provide a collision-free schedule for many contenders, which is desirable in dense scenarios.

	Having a big deterministic backoff increases the time between successful transmissions. Furthermore, if not operating in a scenario with many nodes the empty slots between transmissions are not longer negligible and degrade the overall throughput of the system. For instance, if nodes withdraw from the contention their previously used slots will now be empty. Contenders should be aware of this issue and pursue opportunities to reduce their deterministic backoff, $B_{\text{d}}$ without sacrificing too much in collisions.

	The \emph{Schedule Reset} mechanism for CSMA/ECA$_{\text{Hys+FS}}$ consists on finding the smallest collision-free schedule between a contender's transmissions and then change the node's deterministic backoff to fit in that schedule. Take a contender with $B_{\text{d}}=32$ as an example. By listening to the slots between its transmissions, it is possible to determine the availability of smaller (and possibly) collision-free schedules.

	Figure~\ref{fig:scheduleReset1} shows the slots between the transmissions of a contender with $B_{\text{d}}=32$. Starting from the left, the current $B_{\text{d}}=0$ means that the next slot will be filled with the contender's own transmission. Each following slot containing either a transmission or a collision is identified with the number one, while empty slots are marked with a zero. Notice that the highlighted empty slots appear every eight slots, suggesting that a schedule reduction from $B_{\text{d}}=32$ to $B_{\text{d}}^{*}=8$ is possible\footnote{$B_{\text{d}}=2^{0}CW_{\min}/2=8$ (see Table~\ref{tab:mac-params}). So the change is made to the backoff stage, $k$. In this case from $2$, to $k=0$. This means that any new schedule must also be a power of two.}. The schedule change is performed after the contender's next successful transmission.

	%, coupled with an increase by one in the node's \emph{stickiness}~\cite{L_MAC2}. 
	%\footnote{Trailing waiting periods, like the SIFS, are composed of empty slots. These allow nodes to decrement their backoff counters even after a collision or successful slots. Without loss of generality, the stations in our examples and simulations decrement their respective backoff counters by one after these kind of slots (see example in Figures~\ref{fig:ECA} and~\ref{fig:ecaQoS}).}
		\begin{figure}[tb]
		\centering
			\includegraphics[width=\linewidth]{figures/tonFigs/scheduleReset.pdf}
			\caption{Example of the Schedule Reset mechanism.}
			\label{fig:scheduleReset1}
		\end{figure}

	By default Schedule Reset fills a bitmap $b$ of size $B_{\text{d}}$ with the information observed in each passing slot during $\gamma=B_{\text{d}}/C$ consecutive successful transmissions; where $C=2^{m}CW_{\min}/2$ is the biggest deterministic backoff, and $\gamma$ is referred to as the Schedule Reset \emph{threshold} from this point forward. Each bit $t,~t\in\{0,\ldots ,B_{\text{d}}-1\}$ in the bitmap is the result of a bitwise OR operation between its current value and the state of the corresponding slot. That is, $b[t]\leftarrow b[t]\vee\sigma_{t}(i)$; where $\vee$ is the OR operator and $\sigma_{t}(i)$ is a function that evaluates to $0$ if the slot corresponding to $b[t]$ is empty in iteration $i$, $i \in\{1,\ldots,\gamma\}$; or $1$ otherwise.

	After $i=\gamma$ iterations the bitmap is evaluated using:

	\begin{equation}
		\sum\limits_{t\in A} b[t] =
				\begin{cases}\label{eq:schedEval}
					0, &\text{make change} \\
					>0, &\text{discard}
				\end{cases}
	\end{equation}

	In (\ref{eq:schedEval}), the set $A$ contains all possible shorter schedule lengths, e.g.: $A\in\{2^{j}CW_{\min}/2,\ldots, 2^{k-1}CW_{\min}/2\}$, for $j\in[0,\ldots,k-1]$. Using the example in Figure~\ref{fig:scheduleReset1} to clarify (\ref{eq:schedEval}), we can see that it evaluates to zero for every value of $t\in A$, thus the schedule change is possible.

	The default $\gamma$ ensures that the bitmap registers all the transmission slots in the network (this is true for saturated traffic), providing enough information for performing the schedule reduction without disrupting collision-free schedules. Nevertheless, this $\gamma$ can be rendered too conservative for non-saturated traffic, where contenders enter and leave the contention. This is leveraged by setting $\gamma=1$, meaning that the bitmap is filled with the information of slots between only two consecutive transmissions. Further, there is the option of doing a schedule \emph{halving}. That is, with a set $A\in\{2^{k-1}CW_{\min}/2\}$.

	Schedule Reset can also be configured to increase the node's \emph{stickiness} after a successful reduction of the schedule. Stickiness is not a new concept to CSMA/ECA~\cite{barcelo2011tcf}. It simply instructs the contender to \emph{stick} to the deterministic backoff even in the event of \emph{stickiness} number of consecutive failed transmissions. This allows for a faster convergence towards a collision-free schedule, especially when performing under heavy channel errors.

### 2015-10-08 journal.tex
Completed the section of Schedule Reset

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames.\textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}. Results show that indeed it is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}

>>>>

Beyond simulations results, the implementation of CSMA/ECA prototypes\cite{ECA-DEMO-INFOCOM14, sanabria2013prototyping, BECA-test,CF-MAC} show that the construction of collision-free schedules using a deterministic backoff after successful transmissions is possible and results in a throughput increase by reducing the number of corrupted frames. \textcolor{red}{We then present the first real hardware implementation of CSMA/ECA with Hysteresis using the open firmware OpenFWWF~\cite{OpenFWWF}. Results show that CSMA/ECA is able of providing higher throughput and lower collisions than CSMA/CA for the same number of contenders.}

### 2015-10-08 simulations.tex
Completed the section of Schedule Reset

	\begin{table}
		\centering
		\caption{PHY and MAC parameters for the simulations \textcolor{red}{REVIEW}}
		\label{tab:mac-params}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 65~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Packet size (Bytes) & 1024\\
			MAC queue size (Packets) & 1000\\
			\hline
		\end{tabular}
	\end{table}

>>>>

	\begin{table}
		\centering
		\caption{PHY and MAC parameters for the simulations \textcolor{red}{REVIEW}}
		\label{tab:mac-params}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 65~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Data payload (Bytes) & 1024\\
			MAC queue size (Packets) & 1000\\
			\hline
		\end{tabular}
	\end{table}
